\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-24 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add abstract to JSS paper + minor edits

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

## Example: Applying recursive CV to polynomial regression for the Auto data

@@

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package. Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

>>>>

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function. Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models. The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models. The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

## Example: Applying recursive CV to polynomial regression for the Auto data[^recursive-cv]

[^recursive-cv]: What we call here "recursive CV" has also been termed "nested CV" [e.g., in the \pkg{cvms} package by @OlsenZachariae:2024]. We prefer, however, to reserve the term "nested CV" for the procedure described by [@BatesHastieTibshirani:2023].

@@

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024]. Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

### 2024-05-24 scratch/notes_john/JSS-article/JSS-article-3.tex
add abstract to JSS paper + minor edits

\Abstract{
The abstract of the article.
}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 226376
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.33326
bias-adjusted cross-validation criterion = 19.31477
95% CI for bias-adjusted CV criterion = (15.83621, 22.79333)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
      expr        min         lq       mean     median         uq        max
 hatvalues   1.139062   1.184244   1.202678   1.194248   1.205769   1.297117
  Woodbury  12.106521  12.310701  13.115158  12.337515  12.808810  16.149900
     naive 217.941773 219.491204 225.861284 220.053909 222.015410 278.078974
    cv.glm 373.900074 379.259963 397.907632 382.333036 437.228141 441.112358
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Example: Applying recursive CV to polynomial regression for
the Auto
data}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-data}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x14013a7f0>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package. Here, with minimal explanation, is how one
would perform LOO CV for the example using \pkg{rsample}:\footnote{The
  example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV,
  but otherwise the code from the vignette is only trivially modified
  here.}

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, quietly=TRUE)
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median          uq
      rsample 5814.989369 5864.285227 5952.773502 5897.248509 6023.438043
     cv.exact 4739.827837 4827.064193 4902.546681 4910.249278 4952.683150
      cv.wood  103.663047  104.760125  110.628578  111.201532  116.027007
 cv.hatvalues    1.217905    1.222251    1.409719    1.479649    1.543937
         max neval
 6207.205988    10
 5090.904318    10
  118.610540    10
    1.591538    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to
\code{cv()} with \code{method="exact"}, but more than an order of
magitude slower than \code{cv()} with \code{method="Woodbury"}, and more
than three orders of magnitude slower then \code{cv()} with
\code{method="hatvalues"}.

>>>>

\Abstract{
We describe the \pkg{cv} package, which implements cross-validation for
standard \proglang{R} regression models through a uniform and simple to
use generic function. Methods for the \code{cv()} function are provided
for many commonly employed classes of statistical models, including
mixed-effects models, and it is straightforward to extend the \pkg{cv}
package by writing methods for additional classes of models. The
\code{cv()} function also can cross-validate complex model-selection
procedures, such as those that include variable transformations,
predictor selection, and selection among competing models. The
\code{cv()} function generally supports parallel computations, and the
supplied methods for linear and generalized linear models take advantage
of computational efficiencies.
}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 891621
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.30934
bias-adjusted cross-validation criterion = 19.29207
95% CI for bias-adjusted CV criterion = (15.80161, 22.78253)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min         lq       mean     median         uq        max
 hatvalues    992.282   1149.681   1181.427   1199.188   1223.235   1356.485
  Woodbury  12297.663  12375.317  14678.447  12908.952  16017.675  22227.535
     naive 217204.347 222086.545 249652.690 250411.416 277112.440 279141.530
    cv.glm 379237.823 382052.063 401064.952 388823.726 435052.394 440563.573
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection[Example: Applying recursive CV to polynomial regression for
the Auto data]{\texorpdfstring{Example: Applying recursive CV to
polynomial regression for the Auto
data\footnote{What we call here ``recursive CV'' has also been termed
  ``nested CV'' \citep[e.g., in the \pkg{cvms} package
  by][]{OlsenZachariae:2024}. We prefer, however, to reserve the term
  ``nested CV'' for the procedure described by
  \citep{BatesHastieTibshirani:2023}.}}{Example: Applying recursive CV to polynomial regression for the Auto data}}\label{example-applying-recursive-cv-to-polynomial-regression-for-the-auto-datarecursive-cv}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x1293c2ea8>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package \citep{Kuhn:2024}. Here, with minimal
explanation, is how one would perform LOO CV for this example using
\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
  10-fold rather than LOO CV, but otherwise the code from the vignette
  is only trivially modified here.}

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, quietly=TRUE)
\end{CodeInput}
\begin{CodeOutput}

Attaching package: 'purrr'
\end{CodeOutput}
\begin{CodeOutput}
The following object is masked from 'package:car':

    some
\end{CodeOutput}
\begin{CodeOutput}
The following objects are masked from 'package:foreach':

    accumulate, when
\end{CodeOutput}
\begin{CodeInput}
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median          uq
      rsample 5858.845429 5918.511745 6182.166132 6032.719008 6513.721496
     cv.exact 4823.821011 4879.622175 4986.028516 4973.975331 5051.786505
      cv.wood  103.815690  107.823973  112.389204  112.138341  117.148398
 cv.hatvalues    1.248655    1.299823    1.540665    1.587007    1.654432
         max neval
 6707.589832    10
 5305.047728    10
  121.483287    10
    1.916832    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to
\code{cv()} with \code{method="exact"}, but more than an order of
magnitude slower than \code{cv()} with \code{method="Woodbury"}, and
more than three orders of magnitude slower then \code{cv()} with
\code{method="hatvalues"}.
